Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-07-28 papers/wdmerger_I/paper.tex
Replace references using \autoref from the hyperref package, as advocated by AstroBetter in http://www.astrobetter.com/blog/2014/09/29/latex-hyperref-and-emulateapj/.

@@

This is the first in a series of papers designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. Section \ref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. Section
\ref{sec:initial_models} describes the method we use for setting up a
binary white dwarf simulation. Section \ref{sec:Tests} discusses a few
test problems that we use to demonstrate that our code accurately
solves the equations of fluid dynamics. Section \ref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, Section \ref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

To study the white dwarf merger problem, we use the grid-based
hydrodynamics code CASTRO \citep{castro}. CASTRO solves the Euler
equations, along with the inclusion of optional modules for gravity,
nuclear reactions and thermodynamics. CASTRO is based on the BoxLib
adaptive-mesh refinement (AMR) framework \citep{rendleman:2000}, which
represents fluid data on a mesh where regions of interest have higher
spatial resolution. CASTRO is highly parallel and is designed for
large-scale use on modern supercomputers; see Section
\ref{sec:Performance} for information on how CASTRO performs for our
problem. The next few subsections describe our approach to each of the
physics components used in this work. We direct the reader to the
original code paper for a full description of CASTRO's approach to
solving the equations of hydrodynamics. In this work, we report mainly
on the changes we have made to the code since its original release,
for the purpose of approaching this problem.

@@

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term 
for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form.
The flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum in the domain interior stays constant to
within numerical roundoff error. This result can be derived by analytically recasting 
Equation \ref{eq:momentum_discretization}. In the continuum limit, the two momentum
formulations are identical. Thus the latter has been advocated by, for example, 
\cite{jiang:2013} for the ATHENA code. However, in a finite discretization 
the divergence of the gravitational acceleration is no longer guaranteed to equal
the zone density. In particular, we find that the mixing of the gravitational components
means that the truncation error in the gravitational field can lead to large errors
that imply a density much different than the zone's actual density. This is especially
problematic in a simulation with a low-density ambient medium, where even a small error 
in the momentum update can lead to large changes in a zone's momentum. By continuing to explicitly
use the cell density in the momentum update, we can avoid this possibility: the size of the update
will always be suitably small if the zone's density is small. Thus for our simulations
we continue to use the standard source term for the momentum.

@@

As observed by \cite{arepo}, when examining Equation \ref{eq:mass_transfer} one can see the relationship 
between this method and the more common (non-conservative) approach of evaluating the change in 
gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$ by the gravitational force at the 
cell center. Analytically this form expresses the same core idea via the work-energy theorem, and its 
major flaw is that it evaluates the energy change at the cell center when in fact the mass transfer is 
happening at the cell edges, and the energy change is actually partially compensated by the potential 
in the adjacent zone that the mass is moving to. This is important enough that for simulations where the 
potential can change substantially over the course of the evolution, massive amounts of energy can be lost. 
This standard approach is therefore unacceptable in the case of a problem like white dwarf mergers, and 
the fix to this energy leaking -- evaluating the energy transfer at the six zone faces instead of 
the single zone center -- has a very minor cost in terms of code complexity and computational time.

It is worth pointing out that while Equation \ref{eq:grav_energy_conservation_update} was correctly obtained 
by \cite{arepo} (and indeed that is the inspiration for this approach), there is a subtle error in that derivation. 
Their argument was that if we integrate the work done over the volume of the zone, this volume integral can be 
converted into a surface integral by use of the divergence theorem (see their Equation 93). Then we can simply 
evaluate the work done at each zone face. However, this application of the divergence theorem appears to be 
mathematically incorrect, despite the fact that this is very similar in spirit to the approach discussed here. 
The author fortuitously obtains the correct factor of $1/2$ corresponding to the change in gravitational self-energy.
However, some reflection reveals that there should be no way to derive this double-counting factor simply by integrating 
the work done: that factor is derived from the unique nature of the gravitational potential, which is generated 
by the same distribution that it exerts a force on for a simulation with self-gravity, and it has to be added in 
intentionally to ensure conservation of total energy for these particular types of potentials. We can apply 
the approach of considering energy transfer between the gas energy and the potential energy to any potential field, 
whether or not it is static, changing in time, or generated by the fluid distribution itself. 
Indeed, we will do so for the rotational potential in Section \ref{sec:rotation}, where the 
factor of $1/2$ does not appear. However, it is important also to note that this is not related to a 
global flow of energy or the global nature of the gravitational potential, as claimed by \cite{marcello:2012}. 
This double-counting factor would apply even if there were only two point masses on the grid, since 
work would only be required to bring in the second mass from infinity; the first one can be 
placed at no energy cost. This transfer of energy is explicitly \textit{local} because the 
mass transfer is local. It is true that the mass transfer changes the potential globally, 
but that is implicitly accounted for when using the time-centered potential, and no 
additional steps need to be taken.

Another approach recently taken in the literature for conserving total energy has been to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{marcello:2012} and \cite{jiang:2013}.
This can be done by adding a term corresponding to the ``flux'' of gravitational potential energy. We 
avoid this approach in general because there will be regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the specific implementation of \cite{jiang:2013} 
has terms in their gravitational flux that are not proportional to $\rho$, and so it can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.
(We note in passing too that although \citeauthor{jiang:2013} claim their approach requires a Poisson 
solve for the time derivative of the potential, $\dot{\Phi}$, this is not needed as in practice 
we will already have the potential at time-levels $n$ and $n+1$, and to second order in time, 
$\dot{\Phi}^{n+1/2} = (\Phi^{n+1} - \Phi^{n}) / \Delta t.$)

@@

In a typical simulation we place the boundaries of the domain far
enough away from the region containing most of the mass that some
method of approximation to this direct summation is justified. Many
approaches exist in the literature. The original release of CASTRO
featured the crudest possible approximation: a monopole prescription,
where the boundary values were computed by summing up all the mass on
the domain and treating it as a point source at the domain
center. This is exactly correct only for a spherically symmetric mass
distribution, and therefore is best suited for problems like
single-star Type Ia supernova simulations that employ
self-gravity. However, for a problem like that of a binary star system
with significant departures from spherical symmetry, this assumption
fails to produce accurate boundary values. This results in a
significant drift of the center of the mass of the system over time,
which will be discussed in Section \ref{sec:kepler}.

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = -\sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation
\ref{spherical_harmonic_expansion} using the addition theorem for
spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{align}
  \Phi(\mathbf{x}) &= -\sum_{l=0}^{\infty} \left\{Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} \right. \notag \\
    &+ \left. \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right\}.\label{multipole_potential}
\end{align}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the
outer summation in Equation \ref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, loaded onto the grid by the method described 
in Section \ref{sec:initial_models}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by Equation \ref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we will call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in Figure \ref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. Our choice is also motivated by the 
fact that at this value, the time for computing the boundary conditions on the
coarse grid is comparable to the time for the multigrid solve itself; at higher values 
the boundary condition calculation becomes much more expensive than the multigrid solve
and so the overall returns are severely diminished. We note that in our implementation 
there are no gains at all for going beyond $l_{\text{max}} \gtrsim 16$, as we 
become limited by error due to the machine precision.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

@@

The results of this test are plotted in Figure
\ref{fig:gravity_convergence}. We find that at low resolution 
convergence is actually substantially better than second-order. The 
explanation for this is that we are attempting to model a spherical 
object on a rectangular grid. At very low resolution, the object does 
not look very spherical. As the resolution is increased, the total 
amount of mass on the grid, as well as its location, will change 
as the sphere fills out. This means we are combining the true accuracy 
bonus from increased resolution with the artificial accuracy bonus 
from getting closer to solving the problem we are supposed to be solving. 
On the other hand, at the highest resolution shown the error slightly increases.

We can solve one of these two sources of error by evaluating Equation
\ref{eq:sphere-analytical} with a mass $M$ that corresponds to the
amount of mass actually on the grid, defined as the sum of the density
multiplied by the volume of each cell for all cells whose density is 
greater than ambient. The resolution study for this case is also
plotted in Figure \ref{fig:gravity_convergence}. We still obtain
convergence slightly better than second-order, indicating that we 
still have the geometrical problem of the mass distribution changing.

The only way to fully eliminate this effect is to use a test problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The gravitational potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result by \citet{hummer:1996}, and an earlier calculation 
by \citet{macmillan:1958}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \Phi_{\text{cube}}(x,y,z) &= -G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{Fortran} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and only for this problem 
do we obtain perfect second-order scaling at all resolutions.

>>>>

% allows in-document links; see http://www.astrobetter.com/blog/2014/09/29/latex-hyperref-and-emulateapj/
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref} 
\usepackage[all]{hypcap} %Links go to figures;breaks on deluxetables       
\renewcommand*{\backref}[1]{[#1]}

% Override choices in \autoref
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

@@

This is the first in a series of papers designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
%\hyperref[sec:Numerical Methodology]{Section \ref*{sec:Numerical Methodology}}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:initial_models} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to demonstrate that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

To study the white dwarf merger problem, we use the grid-based
hydrodynamics code CASTRO \citep{castro}. CASTRO solves the Euler
equations, along with the inclusion of optional modules for gravity,
nuclear reactions and thermodynamics. CASTRO is based on the BoxLib
adaptive-mesh refinement (AMR) framework \citep{rendleman:2000}, which
represents fluid data on a mesh where regions of interest have higher
spatial resolution. CASTRO is highly parallel and is designed for
large-scale use on modern supercomputers; see 
\autoref{sec:Performance} for information on how CASTRO performs for our
problem. The next few subsections describe our approach to each of the
physics components used in this work. We direct the reader to the
original code paper for a full description of CASTRO's approach to
solving the equations of hydrodynamics. In this work, we report mainly
on the changes we have made to the code since its original release,
for the purpose of approaching this problem.

@@

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term 
for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form.
The flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum in the domain interior stays constant to
within numerical roundoff error. This result can be derived by analytically recasting 
\autoref{eq:momentum_discretization}. In the continuum limit, the two momentum
formulations are identical. Thus the latter has been advocated by, for example, 
\cite{jiang:2013} for the ATHENA code. However, in a finite discretization 
the divergence of the gravitational acceleration is no longer guaranteed to equal
the zone density. In particular, we find that the mixing of the gravitational components
means that the truncation error in the gravitational field can lead to large errors
that imply a density much different than the zone's actual density. This is especially
problematic in a simulation with a low-density ambient medium, where even a small error 
in the momentum update can lead to large changes in a zone's momentum. By continuing to explicitly
use the cell density in the momentum update, we can avoid this possibility: the size of the update
will always be suitably small if the zone's density is small. Thus for our simulations
we continue to use the standard source term for the momentum.

@@

As observed by \cite{arepo}, when examining \autoref{eq:mass_transfer} one can see the relationship 
between this method and the more common (non-conservative) approach of evaluating the change in 
gas energy using the work done $(\mathbf{v} \cdot \rho \mathbf{g})$ by the gravitational force at the 
cell center. Analytically this form expresses the same core idea via the work-energy theorem, and its 
major flaw is that it evaluates the energy change at the cell center when in fact the mass transfer is 
happening at the cell edges, and the energy change is actually partially compensated by the potential 
in the adjacent zone that the mass is moving to. This is important enough that for simulations where the 
potential can change substantially over the course of the evolution, massive amounts of energy can be lost. 
This standard approach is therefore unacceptable in the case of a problem like white dwarf mergers, and 
the fix to this energy leaking -- evaluating the energy transfer at the six zone faces instead of 
the single zone center -- has a very minor cost in terms of code complexity and computational time.

It is worth pointing out that while \autoref{eq:grav_energy_conservation_update} was correctly obtained 
by \cite{arepo} (and indeed that is the inspiration for this approach), there is a subtle error in that derivation. 
Their argument was that if we integrate the work done over the volume of the zone, this volume integral can be 
converted into a surface integral by use of the divergence theorem (see their Equation 93). Then we can simply 
evaluate the work done at each zone face. However, this application of the divergence theorem appears to be 
mathematically incorrect, despite the fact that this is very similar in spirit to the approach discussed here. 
The author fortuitously obtains the correct factor of $1/2$ corresponding to the change in gravitational self-energy.
However, some reflection reveals that there should be no way to derive this double-counting factor simply by integrating 
the work done: that factor is derived from the unique nature of the gravitational potential, which is generated 
by the same distribution that it exerts a force on for a simulation with self-gravity, and it has to be added in 
intentionally to ensure conservation of total energy for these particular types of potentials. We can apply 
the approach of considering energy transfer between the gas energy and the potential energy to any potential field, 
whether or not it is static, changing in time, or generated by the fluid distribution itself. 
Indeed, we will do so for the rotational potential in \autoref{sec:rotation}, where the 
factor of $1/2$ does not appear. However, it is important also to note that this is not related to a 
global flow of energy or the global nature of the gravitational potential, as claimed by \cite{marcello:2012}. 
This double-counting factor would apply even if there were only two point masses on the grid, since 
work would only be required to bring in the second mass from infinity; the first one can be 
placed at no energy cost. This transfer of energy is explicitly \textit{local} because the 
mass transfer is local. It is true that the mass transfer changes the potential globally, 
but that is implicitly accounted for when using the time-centered potential, and no 
additional steps need to be taken.

Another approach recently taken in the literature for conserving total energy has been to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{marcello:2012} and \cite{jiang:2013}.
This can be done by adding a term corresponding to the ``flux'' of gravitational potential energy. We 
avoid this approach in general because there will be regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the specific implementation of \cite{jiang:2013} 
has terms in their gravitational flux that are not proportional to $\rho$, and so it can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.
(We note in passing too that although \cite{jiang:2013} claim their approach requires a Poisson 
solve for the time derivative of the potential, $\dot{\Phi}$, this is not needed as in practice 
we will already have the potential at time-levels $n$ and $n+1$, and to second order in time, 
$\dot{\Phi}^{n+1/2} = (\Phi^{n+1} - \Phi^{n}) / \Delta t.$)

@@

In a typical simulation we place the boundaries of the domain far
enough away from the region containing most of the mass that some
method of approximation to this direct summation is justified. Many
approaches exist in the literature. The original release of CASTRO
featured the crudest possible approximation: a monopole prescription,
where the boundary values were computed by summing up all the mass on
the domain and treating it as a point source at the domain
center. This is exactly correct only for a spherically symmetric mass
distribution, and therefore is best suited for problems like
single-star Type Ia supernova simulations that employ
self-gravity. However, for a problem like that of a binary star system
with significant departures from spherical symmetry, this assumption
fails to produce accurate boundary values. This results in a
significant drift of the center of the mass of the system over time,
which will be discussed in \autoref{sec:kepler}.

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = -\sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
%% \begin{equation}
%%   q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
%% \end{equation}
After some algebraic simplification of \autoref{spherical_harmonic_expansion} 
using the addition theorem for spherical harmonics,
%% \begin{align}
%%   &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
%%   &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
%% \end{align}
the potential outside of the mass distribution can be written as:
\begin{align}
  \Phi(\mathbf{x}) &= -\sum_{l=0}^{\infty} \left\{Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} \right. \notag \\
    &+ \left. \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right\}.\label{multipole_potential}
\end{align}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
%% \begin{align}
%%   Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
%%   Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
%%   Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
%% \end{align}
This approach becomes computationally feasible when we cut off the
outer summation in \autoref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

As a demonstration of the method's accuracy, we consider the case of two 
white dwarfs of mass ratio 2/3, loaded onto the grid by the method described 
in \autoref{sec:initial_models}. We terminated the simulation just after
initialization, so that we perform only an initial Poisson solve for this 
density distribution. We did this for values of $l_{\text{max}}$ ranging
from 0 to 20, and we also did this using the numerically exact solution 
provided by \autoref{eq:direct_sum}.  Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we computed the $L^2$ error of $\Phi$ on the entire domain for multipole 
boundary conditions, which we will call $\Phi_l$, relative to $\Phi$ 
obtained using the exact boundary conditions:
\begin{equation}
  \text{Error}_l = \frac{\|\Phi_l - \Phi_{\text{exact}}\|_2}{\|\Phi_{\text{exact}}\|_2}.
\end{equation}
The result is shown in \autoref{fig:bc_comparison}. At $l_{\text{max}} = 6$,
the error is already well below $10^{-4}$ and we adopt this as our default 
choice for all simulations with Poisson gravity. Our choice is also motivated by the 
fact that at this value, the time for computing the boundary conditions on the
coarse grid is comparable to the time for the multigrid solve itself; at higher values 
the boundary condition calculation becomes much more expensive than the multigrid solve
and so the overall returns are severely diminished. We note that in our implementation 
there are no gains at all for going beyond $l_{\text{max}} \gtrsim 16$, as we 
become limited by error due to the machine precision.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/bc_comparison}
  \caption{Error of $\Phi$ on the computational domain for a binary white dwarf simulation 
    whose boundary conditions were computed using various values of the maximum multipole order,
    relative to the exact solution determined by a brute force sum on the boundaries.
    Circles represent the error at integer values, and they have been connected by a smooth 
    line to guide the eye.\label{fig:bc_comparison}}
\end{figure}

@@

The results of this test are plotted in \autoref{fig:gravity_convergence}. 
We find that at low resolution 
convergence is actually substantially better than second-order. The 
explanation for this is that we are attempting to model a spherical 
object on a rectangular grid. At very low resolution, the object does 
not look very spherical. As the resolution is increased, the total 
amount of mass on the grid, as well as its location, will change 
as the sphere fills out. This means we are combining the true accuracy 
bonus from increased resolution with the artificial accuracy bonus 
from getting closer to solving the problem we are supposed to be solving. 
On the other hand, at the highest resolution shown the error slightly increases.

We can solve one of these two sources of error by evaluating 
\autoref{eq:sphere-analytical} with a mass $M$ that corresponds to the
amount of mass actually on the grid, defined as the sum of the density
multiplied by the volume of each cell for all cells whose density is 
greater than ambient. The resolution study for this case is also
plotted in \autoref{fig:gravity_convergence}. We still obtain
convergence slightly better than second-order, indicating that we 
still have the geometrical problem of the mass distribution changing.

The only way to fully eliminate this effect is to use a test problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The gravitational potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result by \citet{hummer:1996}, and an earlier calculation 
by \citet{macmillan:1958}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \Phi_{\text{cube}}(x,y,z) &= -G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, 
$y_1 = R - y$, $z_0 = R + z$, $z_1 = R - z$, 
and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{Fortran} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error for this problem is also shown
in \autoref{fig:gravity_convergence}, and only for this problem 
do we obtain perfect second-order scaling at all resolutions.
